---
title: "Google Data Analytics Capstone Project-Bike-Share"
author: "Sakshi"
date: "8/2/2021"
output: html_document
---

## Setting up my R environmnet by installing and loading required packages

```{r}
install.packages("tidyverse")
install.packages("dplyr")
install.packages("lubridate")
install.packages("ggplot2")
library(tidyverse)
library(dplyr)
library(lubridate)
library(ggplot2)
```

## Setting up the working directory where the data files are saved.

```{r}
setwd("/Users/saksh/Desktop/Data Analytics Capstone Project/Data/Unzipped Files/.CSV")
```

# STEP 1: COLLECT DATA
## Loading and renaming all the data files (12 Months) to make easier to work on

```{r}
Jul_2020 <- read.csv("202007-divvy-tripdata.csv")
Aug_2020 <- read.csv("202008-divvy-tripdata.csv")  
Sep_2020 <- read.csv("202009-divvy-tripdata.csv")
Oct_2020 <- read.csv("202010-divvy-tripdata.csv")
Nov_2020 <- read.csv("202011-divvy-tripdata.csv")
Dec_2020 <- read.csv("202012-divvy-tripdata.csv")
Jan_2021 <- read.csv("202101-divvy-tripdata.csv")
Feb_2021 <- read.csv("202102-divvy-tripdata.csv")
Mar_2021 <- read.csv("202103-divvy-tripdata.csv")
Apr_2021 <- read.csv("202104-divvy-tripdata.csv")
May_2021 <- read.csv("202105-divvy-tripdata.csv")
Jun_2021 <- read.csv("202106-divvy-tripdata.csv")
```

# STEP 2 - COMPARE COLUMN NAMES EACH OF THE FILES
## Each file has 13 Column Names but, need to check if they are same in all the files by quering the below functions.

```{r}
colnames(Jul_2020)
colnames(Aug_2020)
colnames(Sep_2020)
colnames(Oct_2020)
colnames(Nov_2020)
colnames(Dec_2020)
colnames(Jan_2021)
colnames(Feb_2021)
colnames(Mar_2021)
colnames(Apr_2021)
colnames(May_2021)
colnames(Jun_2021)
```
## To view and compare data structure
```{r}
str(Jul_2020)
str(Aug_2020)
str(Sep_2020)
str(Oct_2020)
str(Nov_2020)
str(Dec_2020)
str(Jan_2021)
str(Feb_2021)
str(Mar_2021)
str(Apr_2021)
str(May_2021)
str(Jun_2021)
```

# STEP 3 - MAKE CHANGES IF THERE ARE ANY DIFFERENCE IN FILES
## By comparing structure of all the files, I found that start_station_id and end_station_id needs to be converted into character to make field's data type consistent because they are different in other files.

```{r}
Jul_2020 <- mutate(Jul_2020, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
Aug_2020 <- mutate(Aug_2020, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
Sep_2020 <- mutate(Sep_2020, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
Oct_2020 <- mutate(Oct_2020, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
Nov_2020 <- mutate(Nov_2020, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
Dec_2020 <- mutate(Dec_2020, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
Jan_2021 <- mutate(Jan_2021, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
Feb_2021 <- mutate(Feb_2021, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
Mar_2021 <- mutate(Mar_2021, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
Apr_2021 <- mutate(Apr_2021, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
May_2021 <- mutate(May_2021, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
Jun_2021 <- mutate(Jun_2021, start_station_id = as.character(start_station_id), end_station_id = as.character(end_station_id))
```

# STEP 4 - WRANGLE DATA AND COMBINE INTO A SINGLE FILE
## Consolidate all the data into one file.

```{r}
consolidated_data <- bind_rows(Jul_2020, Aug_2020, Sep_2020, Oct_2020, Nov_2020, Dec_2020, Jan_2021, Feb_2021, Mar_2021, Apr_2021, May_2021, Jun_2021)
```

## To view if the consolidated_data is aligned and created correctly, executing following functions to review the data.
## See first 6 rows, check column names, Data Summary, Structure of Data, Rows in data frame & Dimensions of the data frame.
```{r}
head(consolidated_data)
```
```{r}
colnames(consolidated_data)
```
```{r}
summary(consolidated_data)
```
```{r}
str(consolidated_data)
```
```{r}
nrow(consolidated_data)
```
```{r}
dim(consolidated_data)
```


## To do more detailed analysis, 5 columns (Date, Month, Day, Year and day of week) needs to be created.

```{r}
consolidated_data$date <- as.Date(consolidated_data$started_at)
consolidated_data$month <- format(as.Date(consolidated_data$date), "%m")
consolidated_data$day <- format(as.Date(consolidated_data$date), "%d")
consolidated_data$year <- format(as.Date(consolidated_data$date), "%Y")
consolidated_data$day_of_week <- format(as.Date(consolidated_data$date), "%A")
```

## New calculated column to add for analysis like "ride_length" (time difference between started_at and ended_at)

```{r}
consolidated_data$ride_length <- difftime(consolidated_data$ended_at,consolidated_data$started_at)
```

## Inspect the structure of data

```{r}
str(consolidated_data)
```

## Convert "ride_length" from factor to numeric so that we can run calculations on the data

```{r}
consolidated_data$ride_length <- as.numeric(as.character(consolidated_data$ride_length))
```

## Remove "bad" data
## The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative
## We will create a new version of the dataframe (v2) since data is being removed

```{r}
consolidated_data_v2 <- consolidated_data[!(consolidated_data$start_station_name == "HQ QR" | consolidated_data$ride_length<0),]
```

# STEP 5: CONDUCT DESCRIPTIVE ANALYSIS
## Descriptive analysis on ride_length
## Straight Average, midpoint number in the ascending array of ride lengths, longest ride and shortest ride 
```{r}
mean(consolidated_data_v2$ride_length)
```
```{r}
median(consolidated_data_v2$ride_length)
```
```{r}
min(consolidated_data$ride_length)
```
```{r}
max(consolidated_data$ride_length)
```

## Compare members and casual users.
```{r}
aggregate(consolidated_data_v2$ride_length ~ consolidated_data_v2$member_casual, FUN = mean)
aggregate(consolidated_data_v2$ride_length ~ consolidated_data_v2$member_casual, FUN = median)
aggregate(consolidated_data_v2$ride_length ~ consolidated_data_v2$member_casual, FUN = max)
aggregate(consolidated_data_v2$ride_length ~ consolidated_data_v2$member_casual, FUN = min)
```

# See the average ride time by each day for members vs casual users

```{r}
aggregate(consolidated_data_v2$ride_length ~ consolidated_data_v2$member_casual + consolidated_data_v2$day_of_week, FUN = mean)
```

## Notice that the days of the week are out of order. Let's fix that.

```{r}
consolidated_data_v2$day_of_week <- ordered(consolidated_data_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

# Now, let's run the average ride time by each day for members vs casual users

```{r}
aggregate(consolidated_data_v2$ride_length ~ consolidated_data_v2$member_casual + consolidated_data_v2$day_of_week, FUN = mean)
```

## Analyze ridership data by user type and weekday
## This function will show Weekday, Number of rides and average duration by user type.
```{r}
consolidated_data_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%  
  summarise(number_of_rides = n() ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)							
```

# STEP 6 : Disable Scientific Notation of ggplot2 Axis
## The scales package is needed to change the numbers of our plot axes

```{r}
install.packages("scales")
```

```{r}
library(scales)
```

## Let's create visualization by total average duration by user type for last 12 months

```{r}
consolidated_data_v2 %>% 
   group_by(member_casual) %>%
  summarise (average_duration = mean(ride_length)) %>%
  arrange(member_casual) %>%
  ggplot(aes(x = member_casual, y = average_duration , fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_time()
```

## Let's create visualization by no. of rides by user type for last 12 months

```{r}
consolidated_data_v2 %>% 
   group_by(member_casual) %>%
  summarise (number_of_rides = n() ,average_duration = mean(ride_length)) %>%
  arrange(member_casual) %>%
  ggplot(aes(x = member_casual, y = number_of_rides , fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma)
```

## Let's visualize the number of rides by rider type during the week

```{r}
consolidated_data_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n() ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
    geom_col(position = "dodge") +
    scale_y_continuous(labels = comma)
```


## Let's create a visualization for average duration during the week

```{r}
consolidated_data_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n() ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_time()
```

## Let's create a visualization for total number of rides by month

```{r}
consolidated_data_v2 %>% 
  group_by(member_casual, month) %>% 
  summarise(number_of_rides = n() ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, month)  %>% 
  ggplot(aes(x = month , y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_time()
```

## Let's create a visualization for popular bikes by user type

```{r}
consolidated_data_v2 %>% 
  group_by(member_casual, rideable_type) %>%
  summarise(number_of_rides = n()) %>%
  arrange(member_casual, rideable_type) %>%
  ggplot(aes(x = rideable_type, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma)
```
